Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-15 Paper/Draft_harmonization_paper.tex
15.02: Added author contributions and keywords

% In press or forthcoming: van Bergen, E., de Jong, P. F., Maassen, B., Krikhaar, E., Plakas, A., & van der Leij, A. (in press). IQ of four-year-olds who go on to develop dyslexia. Journal of Learning Disabilities. DOI: 10.1177/0022219413479673
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Leon Metzeling}
\author[4]{Astrid Schmidt-Kloiber}
\author[4]{Wolfram Graf}
\author[5]{Philippe Usseglio-Polatera}
\author[6]{Christoph D. Matthaei}
\author[7]{Charles P. Hawkins}
\author[8]{N. LeRoy Poff}
\author[9]{Laura Twardochleb}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[4]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[7]{Department of Watershed Sciences and the Ecology Center, Utah State University, Logan, USA}
\affil[8]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[9]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\date{}

@@

4. By outlining discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.

@@

The heterogeneity of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of studies across continents. There are several challenges presented by using invertebrate trait data from multiple established databases in ecological studies. One such challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, where a grouping feature is defined as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species". Thus, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

@@

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) when comparing among trait databases, or when linking observational taxonomic data to trait databases, presents another challenge. When observations or data in one trait database are at a more precise taxonomic level than other data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available at more precise taxonomic levels than observed taxa, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing. 
%However, related knowledge would inform future studies regarding the choice of the aggregation method.

@@

Harmonised invertebrate trait datasets for Australia, Europe, North America and New Zealand are deposited \textit{Zenondo/Github?}

@@

% Verena Schreiner body form traits?
We thank Brooke Cassell for technical editing of the manuscript. The project is funded by the German research society (DFG \textit{insert project number}).

@@

>>>>

% In press or forthcoming: van Bergen, E., de Jong, P. F., Maassen, B., Krikhaar, E., Plakas, A., & van der Leij, A. (in press). IQ of four-year-olds who go on to develop dyslexia. Journal of Learning Disabilities. DOI: 10.1177/0022219413479673
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%% New Commands %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Keywords command
\providecommand{\keywords}[1]
{
  \small	
  \textbf{\textit{Keywords---}} #1
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution}
\author[1]{Stefan Kunz}
\author[2]{Ben J. Kefford}
\author[3]{Astrid Schmidt-Kloiber}
\author[4]{Christoph D. Matthaei}
\author[5]{Philippe Usseglio-Polatera}
\author[4]{Wolfram Graf}
\author[6]{N. LeRoy Poff}
\author[7]{Leon Metzeling}
\author[8]{Laura Twardochleb}
\author[9]{Charles P. Hawkins}
\author[1]{Ralf B. Schäfer}
\affil[1]{Institute for Environmental Sciences, University of Koblenz-Landau, Landau, Germany}
\affil[2]{Centre for Applied Water Science, Institute for Applied Ecology, University of Canberra, Canberra, Australia}
\affil[3]{Institute of Hydrobiology and Aquatic Ecosystem Management, University of Natural Resources and Life Sciences Vienna (BOKU), Vienna, Austria}
\affil[4]{Department of Zoology, University of Otago, Dunedin, New Zealand}
\affil[5]{University of Lorraine, CNRS, LIEC, Metz, France}
\affil[6]{Department of Biology, Colorado State University, Fort Collins, USA}
\affil[7]{Environment Protection Authority Victoria, Applied Sciences Division, Macleod, Australia}
\affil[8]{Department of Fisheries and Wildlife, Michigan State University, East Lansing, USA}
\affil[9]{Department of Watershed Sciences and the Ecology Center, Utah State University, Logan, USA}
\date{}

@@

4. By outlining discrepancies in trait definitions we hope to motivate the development of standardised terminology for invertebrate traits. Our results also illustrate the usefulness of harmonised datasets for ecological study and provide guidance for the circumstances under which the choice of trait aggregation method is important.
\\
\\
\keywords{trait definitions, trait aggregation, data synthesis}

@@

The heterogeneity of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of studies across continents. There are several challenges presented by using invertebrate trait data from multiple established databases in ecological studies. One such challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, where a grouping feature is defined as a general property (e.g. feeding mode) that comprises a "group of related traits (e.g., predator, shredder, etc.) that vary among species or among individuals within a species" (page number 825). Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

@@

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) when comparing among trait databases, or when linking observational taxonomic data to trait databases, presents another challenge. When observations or data in one trait database are at a more precise taxonomic level than other data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). Conversely, if trait information is only available at more precise taxonomic levels than observed taxa, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

@@

Harmonised invertebrate trait datasets for Australia, Europe, North America and New Zealand are deposited on Github. % Add link

@@

% Verena Schreiner body form traits?
We thank Brooke Cassell for technical editing of the manuscript. The project, in which this study was conducted, is funded by the German research society (DFG, project number 338785727).

@@

\section*{Author contribution}
R.B.S. and S.K. designed the study; B.J.K. and L.M. provided trait data for Australia; P.U-P. provided trait data from the Tachet database; C.P.H., N.L.P., and L.T. provided trait data for North America; C.D.M. provided trait data for New Zealand;
B.J.K., A.S-K., C.D.M, W.G., N.L.P., L.M., and C.P.H. provided in depth information on trait data and discussed procedure of analysis; S.K. conducted data processing and the statistical analyses.; S.K. wrote the manuscript; B.J.K, A-S.K., C.D.M, P.U-P., C.P.H., and R.B.S. discussed the results and provided suggestions on the manuscript.
